Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2023-10-30 manuscript/wp4.tex
ms 27/10

\PassOptionsToPackage{hidelinks}{hyperref}
\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
\usepackage{amsmath}
\usepackage{mathspec}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}

\title{Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography}

\shorttitle{The cultural evolution of informative orthography}

\authorsnames[1,1]{Jon W. Carr, Kathleen Rastle}

@@

% \authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2023). Informativity in the writing system. \textit{OSF Preprints}. Version~0.}

\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to acquire fully. It is less widely recognized, however, that such orthographies may offer certain benefits in the context of reading. For example, the heterographic homophones (e.g., <knight> vs. <night>) impose additional costs in learning but reduce ambiguity in reading. Could it be the case, then, that heterography---and therefore orthographic opacity---may sometimes be selected for in the evolution of writing systems, despite the cost in learning? We investigate this question by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity. In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises through exaptation as an epiphenomenon of sound change (e.g., <meat> vs. <meet> resulting from the /ɛː/–/eː/ merger in the Great Vowel Shift). Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, these systems can become informative, diverging from the spoken language in order to express meaning directly.}

@@

\noindent
Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each written letter corresponding to a distinct sound. However, writing systems are known to diverge from their spoken counterparts. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech. It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading.

One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings (e.g., <meat> and <meet>, both pronounced /miːt/). Heterographic spellings such as these can serve a valuable function in reading. For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on parsing the upcoming syntax. Heterography may, therefore, confer certain benefits in reading, permitting a writing system to be---at least in some areas---more informative than its spoken counterpart. Although such divergence is common in the more orthographically opaque languages, such as English, even the most transparent of languages have at least some techniques for disambiguating homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).

% The same is true of, for example, the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).

% Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages. For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis. The writing system, however, does not simply replace the <p> with a <b>, as one might predict; instead, both letters are included, allowing the orthography to, at~once, show meaning and phonetic form: the eclipsed spelling of <páistí> is <bpáistí>. Moreover, in some cases, the eclipsed spellings result in ambiguity avoidance---in this case, if the <p> were not retained, the hypothetical spelling <báistí> would be confusable with the genitive form of the word for rain.

However, the pursuit of additional informativeness in orthography comes at a cost. A system of words that conveys more information will---in general---be more complex than one that conveys less information. There exists a tradeoff between the simplicity of the spelling system and its ability to be expressive. A simple spelling system is one that is easy to learn, use, and process; for example, by being transparent with respect to phonology. An informative spelling system, on the other hand, is one that precisely conveys meaning. The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in a wide range of both typological \parencite{KempRegier:2012, Kemp:2018} and experimental \parencite{Kirby:2008, Kirby:2015} studies. Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication. An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness in use that is outweighed by the cost of learning arbitrary spelling distinctions.

% There seem to be broadly two kinds of benefit. (1) disambiguation in the writing of words that would otherwise be ambiguous if spelled transparently. (2) permitting the cognitive system to rapidly and accurately attain the meaning of a word. In this paper we limit ourselves to informativeness of the first kind, which is more amenable to experimental testing.

This leads to the intriguing question of whether heterography might be selected for in the cultural evolution of orthography and, if so, the mechanisms through which this selection could occur. \textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms. For example, the word \textit{lite} is a respelling of \textit{light} frequently used in food products to distinguish light-in-calorific-weight from other meanings of \textit{light}; in British English, the word \textit{cheque} (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc. Heterography is also commonly employed in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}).

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two (or more) spoken forms merge and become homophonous, but the original heterographic spellings are conserved in the orthography. In English, for example, the meat--meet merger ultimately resulted in Middle English /ɛ:/ (spelled <ea>) and /e:/ (spelled <ee>) merging into /i:/ in Early Modern English, although the spellings were never updated accordingly, thus giving rise to a set heterographic homophones that persist in present-day English: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}, etc. Similarly, other historic sound changes have likely resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (\textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}, etc.), the loss of /ç/ (\textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}, etc.), and the merger of /ʍ/ into /w/ (\textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}, etc.).

In this paper, we experimentally simulate the processes of differentiation and conservation using the experimental iterated learning paradigm \parencite{Kirby:2008}. In this paradigm, an artificially constructed language is passed along a \textit{transmission chain} of human participants, simulating what happens during the cultural transmission and evolution of language. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding emergent linguistic phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015, Beckner:2017}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, and regularization \parencite{Smith:2010, Ferdinand:2019}. For reviews, see \textcite{Tamariz:2017, Kirby:2014, ScottPhillips:2010}.

One important component of the iterated learning paradigm is the \textit{bottleneck on transmission}, a limit on the amount of information flowing from one generation to the next. Under a tight bottleneck, where little data passes from one generation to the next, the learner must perform more generalization from less input, such that the cognitive biases that the learner brings to the table become more important in shaping the structure of the emergent language. An important aspect of our studies is how learners respond to this bottleneck---how they choose to generalize.

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that the heterographic representation of meanings expressed homophonously in speech arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io}. Both experiments were preregistered at \url{https://aspredicted.org/YLG_CKD}.

@@

Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography. In particular we had two main hypotheses:

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. Each participant is first asked to learn a simple ``alien'' language---consisting of words for colored shapes---and is then asked to reproduce what they learned in a test phase. To explore the hypotheses outlined above, we conducted the experiment under two different conditions: Transmission-only, in which the test phases emphasizes simple reproduction, and Transmission~+ Communication, in which the test phase encourages disambiguation.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
	\vspace*{2pt}
	\caption{The nine alien object stimuli with their stems and suffixes. The spoken and written forms of the stems were fixed and unchanging throughout the experiment, as were the spoken forms of the suffixes, which were always homophonous; however, the written forms of the suffixes were free to evolve over time, potentially taking on differentiated forms to indicate color (e.g., <-co>, <-ko>, and <-qo> to represent pink, yellow, and blue).}
	\label{stimuli}
	\end{figure}

Participants were taught words for nine ``alien'' objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.

Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system, which was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>. In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity. This procedure models an initial state of high spelling variation (every object has a unique suffix spelling), but these spellings may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning). The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{(A) Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. (B) Transmission~+ Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}

>>>>

\PassOptionsToPackage{hidelinks}{hyperref}

\documentclass[doc,biblatex]{apa7}

\usepackage{amsmath}
\usepackage{mathspec}

\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}

\captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}

\title{Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography}

\shorttitle{Why do languages tolerate heterography?}

% PREPRINT VERSION
\authorsnames{Jon W. Carr, Kathleen Rastle}

@@

\authornote{\textbf{Cite as:} Carr, J.\@ W., \& Rastle, K.\@ (2023). Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography. \textit{OSF Preprints}. Version~1. https://doi.org/10.31219/osf.io/r5n3g}

% SUBMISSION VERSION
% \authorsnames{Anonymous}

% \authorsaffiliations{{Anonymous}}

\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to acquire fully. It is less widely recognized, however, that such orthographies may offer certain benefits in the context of reading. For example, heterographic homophones (e.g., <knight> vs. <night>) impose additional costs in learning but reduce ambiguity in reading. If correct, this would suggest that heterography---and therefore orthographic opacity---might sometimes be selected for in the evolution of writing systems, despite the cost in learning. We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity. In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> vs. <meet> resulting from the /ɛː/–/eː/ merger in the Great Vowel Shift). Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, these systems can become informative, diverging from the spoken language in order to express meaning directly.}

@@

\noindent
Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each letter corresponding to a distinct sound. However, writing systems are known to diverge from their spoken counterparts. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}. It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading \parencite{Ulicheva:2020}.

One potential case of functional divergence is heterographic homophones---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}). Heterographic spellings such as these may serve a valuable function in reading. For example, faced with a spoken sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr kat/, a form of the verb \textit{to~be}, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content. Heterography permits the written form of the language to be more informative than the spoken form. Although heterography is common in the more orthographically opaque languages, such as English, Hebrew, and Thai, even the most transparent of languages employ at~least some techniques to disambiguate homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).

Heterographic homophony can also occur at the morpheme level. The suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing. Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English morphemes \parencite{Berg:2017} and it has been shown that English readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}.

However, heterography comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. The heterographic spellings <meat> and <meet>, for example, imply that the /i/ sound can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, making alphabetic decoding more difficult \parencite{Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}. Secondly, the mapping between heterographic forms and meaning must also be learned. From the point of view of the modern language user, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>; it could just as easily be the other way around. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Shankweiler:1992}.

An adaptive explanation for heterographic homophones must, then, demonstrate that heterography confers a benefit in terms of informativeness that outweighs its associated costs. In general, a system of words that conveys more information will be more complex than one that conveys less information; there exists a tradeoff between the simplicity of the spelling system and its ability to be expressive. A simple spelling system would be one that is easy to learn, use, and process; for example, by being transparent with respect to phonology. An informative spelling system, on the other hand, would be one that precisely conveys meaning. The idea of a tradeoff between simplicity and informativeness in language structure has a longstanding history \parencite{Gabelentz:1891, Zipf:1949, Martinet:1952, Rosch:1978} and has recently been explored in various typological \parencite[e.g.,][]{Kemp:2018} and experimental \parencite[e.g.,][]{Kirby:2015} studies. Of particular note here is the idea that the pressure for simplicity in language structure derives from learning, while the pressure for informativeness derives from the need for precise communication, a point we return to shortly.

First, however, it is useful to consider the mechanisms by which selection could occur if it is indeed the case that heterography emerges for functional reasons. \textcite[pp.~325--326]{Berg:2021} outline two models of how a word might enter a state of heterography. The first model, \textit{the differentiation model} (Fig.~\ref{models_of_heterography}A), explains heterography through the creation of new orthographic forms. For example, the spelling <lite> for the word \textit{light} is frequently used in food products to mean light-in-calorific-weight; in British English, the spelling <cheque> (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Many monosyllabic words that are homophonous with common function words also tend to adopt alternate spellings, often by adding <e> or by doubling the final consonant: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, \textit{we}--\textit{wee}. Differentiation is also common in surnames---\textit{Clarke}, \textit{Greene}, \textit{Wilde}; \textit{Carr}, \textit{Hogg}, \textit{Mann}--- and trade names---\textit{Blu Tack}, \textit{Froot Loops}, \textit{Wite-Out} \parencite[§~6]{Carney:1994}. One important way in which differentiation can occur---especially in a language like English that has historically contained a lot of spelling variation \parencite{Nevalainen:2012, Stenroos:2016}---is by the conditioning of variant spellings on meaning; pairs like \textit{discreet}--\textit{discrete}, \textit{flour}--\textit{flower}, and \textit{plain}--\textit{plane}, which were once variant spellings of the same word, have taken on distinct meanings over time \parencite[§~5.4]{Carney:1994}. \textcite[p.~58]{Berg:2017} have referred to this as the ``functionalization of leftovers'': The spelling variants that survive are those that ``can find distributional or functional niches.''

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. \textbf{(A)} In the differentiation model, two meanings are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. \textbf{(B)} In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

The second model, \textit{the conservation model} (Fig.~\ref{models_of_heterography}B), explains heterographic homophones as the historical residue of sound change: Two spoken forms merge and become homophonous, but the original spellings are conserved in the orthography. For example, the \textit{meat}--\textit{meet} merger that occurred during the Great Vowel Shift ultimately resulted in Middle English /ɛː/ (spelled <ea>) and /eː/ (spelled <ee>) being pronounced /iː/ in Early Modern English \parencite{Lass:2000}, but the spellings were never changed accordingly, thus giving rise to a set of heterographic homophones that persist in present-day English \parencite[pp.~140--141]{Wells:1982}: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}. The same is true of the \textit{pain}--\textit{pane} merger \parencite[pp.~141--142]{Wells:1982}: \textit{maid}--\textit{made}, \textit{main}--\textit{mane}, \textit{pain}--\textit{pane}, \textit{raise}--\textit{raze}, \textit{sail}--\textit{sale}, \textit{vain}--\textit{vane}. Sound changes involving consonants have also resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (e.g., \textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}), the loss of /ç/ (e.g., \textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}), and the merger of /ʍ/ into /w/ (e.g., \textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}).\footnote{Although we can never be entirely certain how words were pronounced before the advent of sound recording technology, historical linguists have compiled persuasive evidence by a variety of methods. Comparison to modern Dutch and German, for example, offers an insight into how these words might have been pronounced in the past (e.g., \textit{knot} is cognate with German \textit{Knoten} where the /kn/ cluster continues to be fully rendered and \textit{eight} is cognate with Dutch and German \textit{acht}, where the palatal fricative still exists).} A more recent (and perhaps in-progress) example can be found in dialects that have undergone the \textit{father}--\textit{bother} merger, including most varieties of American English \parencite[p.~169]{Labov:2005}, which has resulted in, for example, \textit{balm}--\textit{bomb} (/bɑm/), \textit{lager}--\textit{logger} (/lɑɡər/), and \textit{mach}--\textit{mock} (/mɑk/); these words continue to be spelled with <a> vs.\@ <o> despite being homophonous in such dialects.

In some cases, it is debatable whether a given case of heterography was delivered by the differentiation or conservation mechanism. For example, while the etymological (and folk-etymological) respellings introduced during the Renaissance might appear to be cases of conservation \parencite[the most notorious example being the replacement of <dout> with <doubt> to indicate the word's Latin derivation from \textit{dubitare};][p.~268]{Crystal:2005}, it has also been argued that such respellings were motivated in~part by a desire to differentiate homophones such as \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, and \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}. Nevertheless, regardless of the particular mechanism behind specific cases in English or any other language, our primary contention here is that both of these mechanisms provide adaptive, functional explanations for heterography. Differentiated spellings that prove communicatively useful will be more likely to survive; likewise, conserved spellings that prove communicatively useful will be more likely to survive.

Our aims in this paper are twofold. First, we test the idea that heterography emerges in response to a functional pressure to disambiguate meaning in writing. Second, we seek to understand how the emergence of heterography plays~out under the two mechanisms of differentiation and conservation. Is one of these a better candidate explanation than the other? Approaching these evolutionary questions using data from natural languages is challenging. In particular, the available diachronic data (for any language) will necessarily be limited and impoverished---languages do not fossilize well, especially in their spoken forms. In addition, any answer derived from such datasets will have to rely on correlational, as opposed to causal, evidence---we cannot rerun history many times under different conditions.

We therefore turn to a different approach. Here we experimentally simulate the processes of differentiation and conservation using the experimental iterated learning paradigm \parencite{Kirby:2008}. In this paradigm, an artificially constructed language (or so-called ``alien language'') is passed along a \textit{transmission chain} of human participants, simulating what happens during the cultural transmission and evolution of language. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and, subsequently, produces new linguistic output for participant $i+1$ to learn from, although the participants themselves are not aware of this generational structure. It has been demonstrated in a wide variety of studies that, after several generations of cultural transmission, artificial languages can gradually adapt to the biases of the human learners and the environments in which they are used, yielding emergent linguistic phenomena, such as compositionality \parencite{Beckner:2017, Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, semantic category structure \parencite{Carr:2017, Canini:2014, Silvey:2019}, regularization \parencite{Ferdinand:2019, Smith:2010}, and argument marking \parencite{Motamedi:2021}, among many other things. For reviews, see \textcite{Bailes:2023, Kirby:2014, Kirby:2017, Smith:2022, Tamariz:2017}.

\textcite{Kirby:2008} described the first experimental application of the iterated learning framework (which had previously been confined to computational modeling), showing that compositional structure---a systematic relationship between recombinant linguistic units and meaning---could spontaneously emerge under a \textit{bottleneck on transmission}. This ``bottleneck'' defines a limit on the amount of information that can flow from one generation to the next \parencite{Brighton:2002}. Under a tight bottleneck, where little data passes from one generation to the next, the learner must perform more generalization from less input, such that the cognitive biases that the learner brings to the table become more important in shaping the structure of the emergent language. Generalization from limited input is a big driver behind systematic structure in the language as a whole, since human learners tend to generalize in ways that increase the simplicity (i.e., systematicity) of the system, albeit unconsciously \parencite{Culbertson:2016}. However, left unchecked, this bias for simplicity would ultimately result in the emergence of maximally simple, degenerate languages. \textcite{Kirby:2015} therefore extended the framework by including a communicative task; instead of each generation consisting of a single participant, each generation now consisted of a pair of participants engaged in a shared task requiring communicative precision. Crucially, this communicative component prevented the artificial languages from degenerating; instead, the languages find a tradeoff between simplicity on the one hand and informativeness on the other, just as in natural language \parencite{Mollica:2021, KempRegier:2012, Kemp:2018, Regier:2015, Zaslavsky:2018}. We follow this experimental approach in order to compare what happens to orthography when a pressure for informativeness \textit{is} and \textit{is~not} present. [KATHY: I feel that someone could wrongly think that we're just replicating what's already been done but with spelling.  Is there a way to remind readers here that we're looking at what pressures a spelling system to diverge from spoken language?  the key in our experiment is that diverging from spoken language increases pressure on learning.]

Our paper reports the results of two experiments---focusing on the differentiation and conservation models respectively---with the goal of demonstrating that functional heterography arises preferentially under a communicative need for disambiguation. All data and code is available from \url{https://osf.io}. Both experiments were preregistered at \url{https://aspredicted.org/YDV_MXV}. Although we did not specify a precise statistical protocol (due to the more exploratory nature of our experiments), our preregistration does cover hypotheses, conditions, measured variables, sample size, and exclusion criteria.

@@

Our first experiment tests the ability of the differentiation model to explain the emergence of informativeness in orthography [KATHY: Possible to remind readers that we mean divergence from spoken language i.e. making the orthography more complex?] (i.e., in the face of homophony). We had two main hypotheses:
\begin{enumerate}
	\item Under pressure from learning alone, we expect to see the emergence of an increasingly transparent orthography.
	\item Under additional pressure for disambiguation, we expect to see greater use of differentiated, non-transparent spellings.
\end{enumerate}

@@

Our methods follow the experimental iterated learning literature, as described above, with one main difference: The artificial language has both spoken and written forms that may diverge or converge over time. Broadly, participants are first asked to learn a simple alien language (consisting of words for colored shapes) and are then asked to reproduce what they learned in a test phase. The written form of the language may change over time, since the orthographic output of participant $i$ becomes the input to participant $i+1$ in a transmission chain design, but the spoken form of the language remains fixed and under experimenter control. To explore the hypotheses outlined above, we conducted the experiment under two different conditions: Transmission-only, in which the test phases emphasizes simple reproduction, and Transmission~+ Communication, in which the test phase encourages disambiguation.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
	\vspace*{2pt}
	\caption{The nine object stimuli with their stems and suffixes. The spoken and written forms of the stems were fixed and unchanging throughout the experiment, as were the spoken forms of the suffixes, which were always homophonous; however, the written forms of the suffixes were free to evolve over time, potentially taking on differentiated forms to indicate color (e.g., <-co>, <-ko>, and <-qo> to represent pink, yellow, and blue).}
	\label{stimuli}
	\end{figure}

Participants were taught words for nine alien objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.

Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system, which was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>. In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contained no particular regularity. This procedure models an initial state of high spelling variation (every object has a unique suffix spelling), but these spellings may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning). The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{\textbf{(A)} Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. \textbf{(B)} Transmission~+ Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}
